---
title: 'Final Project: Tempmerature Data in the World'
output:
  html_document:
    df_print: paged
---

Initial Setup

```{r}
# library all necesssary packages
library(maps)
library(tidyverse)
library(caret)
library(caTools)
library(lubridate)
library(rworldmap)

# set up with working directory then assign datasets to variables
setwd("C:/Users/odettweiler/OneDrive - Eastside Preparatory School/Documents/07 - Data Science/R files/temperature data/")
global_temps <- read.csv("GlobalTemperatures.csv")
temps_by_country <- read.csv("GlobalLandTemperaturesByCountry.csv")
temps_by_city <- read.csv("GlobalLandTemperaturesByMajorCity.csv")

# check it out
summary(temps_by_country)
```

Now to clean and augment the data!

```{r}
# as.Date code from here: https://www.statology.org/as-date-function-in-r/

# add major city data here, cleaning NAs, adding fahrenheit, and converting dates to year

temps_by_country <- temps_by_country |>
  filter(!is.na(AverageTemperature)) |> # removing rows with NA values for temperature
  mutate(AverageTemperatureF = 32 + ((9/5) * AverageTemperature)) |> # add fahrenheit data
  mutate(Date = as.Date(dt)) |> # convert dates to Date type (is this necessary??)
  mutate(Year = strtoi(substr(dt, 1, 4))) |> # add year column for easy filtering
  filter(Year >= 1900) # filter to relevant data

temps_by_city <- temps_by_city |>
  filter(!is.na(AverageTemperature)) |> # removing rows with NA values for temperature
  mutate(AverageTemperatureF = 32 + ((9/5) * AverageTemperature)) |> # add fahrenheit data
  mutate(Date = as.Date(dt)) |> # convert dates to Date type
  mutate(Year = strtoi(substr(dt, 1, 4))) |> # add year column for easy filtering
  mutate(LatNumeric  = as.double(str_sub(Latitude, 0, -2))) |>
  filter(Year >= 1900) # filter to relevant data

# only run once
global_temps <- global_temps |>
  select(-c(LandMaxTemperature, LandMaxTemperatureUncertainty, LandMinTemperature, LandMinTemperatureUncertainty))

global_temps <- global_temps |>
  filter(!is.na(LandAverageTemperature) & !is.na(LandAndOceanAverageTemperature)) |> # removing rows with NA values for temperature
  mutate(LandAverageTemperatureF = 32 + ((9/5) * LandAverageTemperature)) |> # add fahrenheit data
  mutate(LandAndOceanAverageTemperatureF = 32 + ((9/5) * LandAndOceanAverageTemperature)) |>
  mutate(Date = as.Date(dt)) |> # convert dates to Date type
  mutate(Year = strtoi(substr(dt, 1, 4))) |> # add year column for easy filtering
  filter(Year >= 1900) # filter to relevant data

#view(temps_by_city)
#view(temps_by_country)
#view(global_temps)

head(global_temps)
```


Now let's look at some of the data. For countries' data, we'll look at the US, UK, and Brazil. For cities, we'll glimpse at the data for some of the highest points. And finally, we'll take a glance at average global temperatures. 

```{r}

# graphing different countries' data using facet
# code from here: http://www.cookbook-r.com/Graphs/Facets_(ggplot2)/
temps_by_country |>
  filter(Country == "United States" | Country == "United Kingdom" | Country == "Brazil") |>
  ggplot(aes(x = Date, y = AverageTemperatureF)) + geom_point() + geom_smooth() + facet_grid(Country ~ .) + 
  labs(x = "Year", y = "Average Monthly Temperature (F)", title = "Average Monthly Temperature over Time in Different Countries")

temps_by_city |>
  group_by(City, Country, LatNumeric) |>
  summarize(TotalAverageF = mean(AverageTemperatureF)) |>
  filter(TotalAverageF == max(TotalAverageF)) |>
  ungroup() |>
  filter(TotalAverageF == max(TotalAverageF))

global_temps |>
  ggplot(aes(x = Date, y = LandAndOceanAverageTemperatureF)) + geom_point() + geom_smooth() + 
  labs(x = "Year", y = "Global Average Temperature (F)", title = "Global Average Temperature over Time")
```

We're going to filter average this data for each year to make it easier to look at.

```{r}
# country data

# get yearly average
temps_by_country_y <- temps_by_country |>
  group_by(Year, Country) |>
  summarize(YearAvgF = mean(AverageTemperatureF))

# add country codes
temps_by_country_y$ISOName <- lapply(temps_by_country_y$Country, rwmGetISO3)

temps_by_country_y <- temps_by_country_y |> filter(!is.na(ISOName))

temps_by_country_y |>
  filter(Country == "United States" | Country == "United Kingdom" | Country == "Brazil") |>
  ggplot(aes(x = Year, y = YearAvgF)) + geom_point() + geom_smooth() + facet_grid(Country ~ .) + 
  labs(x = "Year", "Yearly Average Temperature (F)", title = "Yearly Average Temperature by Country")

# city data
temps_by_city_y <- temps_by_city |>
  group_by(Year, City, LatNumeric) |>
  summarize(YearAvgF = mean(AverageTemperatureF)) |>
  ungroup()

temps_by_city_y |>
  filter(Year == 2013) |>
  ungroup() |>
  filter(YearAvgF == max(YearAvgF))

# global data
global_temps_y <- global_temps |>
  group_by(Year) |>
  summarize(LandYearAvgF = mean(LandAverageTemperatureF), LandAndOceanYearAvgF = mean(LandAndOceanAverageTemperatureF))

global_temps_y |> # this has a cool visual with a conclusion!
  ggplot(aes(x = Year, y = LandAndOceanYearAvgF)) + geom_point() + geom_smooth() + 
  labs(x = "Year", y = "Yearly Average Global Temperature (F)", title = "Yearly Average Global Temperatures over Time")
```


```{r}
# code from here: https://stackoverflow.com/questions/11225343/how-to-create-a-world-map-in-r-with-specific-countries-filled-in
map_colors <- c('lightblue', 'yellow', 'orange', 'red')

heatMapDF_recent <- temps_by_country_y |> ungroup() |> filter(Year == 2013) |> select(ISOName, YearAvgF)
heatMapDF_old <- temps_by_country_y |> ungroup() |> filter(Year == 1913) |> select(ISOName, YearAvgF)
# malDF is a data.frame with the ISO3 country names plus a variable to
# merge to the map data

heatMap_recent <- joinCountryData2Map(heatMapDF_recent, joinCode = "ISO3",
  nameJoinColumn = "ISOName")
heatMap_old <- joinCountryData2Map(heatMapDF_old, joinCode = "ISO3",
  nameJoinColumn = "ISOName")
# This will join your malDF data.frame to the country map data

#heatMap

mapCountryData(heatMap_recent, nameColumnToPlot="YearAvgF", catMethod = "pretty", colourPalette = map_colors,
  missingCountryCol = gray(.8), addLegend = TRUE, borderCol = "black", mapTitle = "Average Yearly Temperature by Country (2013)")
mapCountryData(heatMap_old, nameColumnToPlot="YearAvgF", catMethod = "pretty", colourPalette = map_colors,
  missingCountryCol = gray(.8), addLegend = TRUE, borderCol = "black", mapTitle = "Average Yearly Temperature by Country (1913)")
# And this will plot it, with the trick that the color palette's first
# color is red


# more code from here: https://www.rdocumentation.org/packages/rworldmap/versions/1.3-6
```





```{r}
# make a linear model with time to predict overall temperatures in the future

cor(global_temps_y$Year, global_temps_y$LandAndOceanYearAvgF)
# nice correlation, let's look for outliers

ggplot(global_temps_y) + geom_boxplot(aes(y = LandAndOceanYearAvgF)) + labs(y = "Global Average Temperatures (F)")
# not too bad on outliers, let's keep going

# look for normality
ggplot(global_temps_y) + geom_histogram(aes(x = LandAndOceanYearAvgF)) + labs(x = "Global Yearly Averages", y = "Count", title = "Global Yearly Averages")
# pretty normal, let's make the linear model now
```
Our data looks pretty good, let's actually do the thing. 

```{r}
global_lm <- lm(LandAndOceanYearAvgF ~ Year, data = global_temps_y)
summary(global_lm)
```
Now a plot with the fit line:

```{r}
eq = paste0("Global Yearly Averages vs. Temperature Model: y = ", round(global_lm$coefficients[2],2),
            "*x + ", round(global_lm$coefficients[1],2))

global_temps_y |>
  ggplot(aes(x = Year, y = LandAndOceanYearAvgF)) + geom_point(size = 2) +
  xlab("Year") +
  ylab("Global Yearly Averages") +
  # add an arbitrary line with slope and intercept
  geom_abline(intercept = global_lm$coefficients[1], # extract first coefficient from model
              slope = global_lm$coefficients[2], # ditto
              linetype = "dashed",
              color = "blue",
              linewidth = 1.5) +
  ggtitle(eq) +
  theme(plot.title = element_text(color="blue", size=14, face="bold"))
```
```{r}
predict(global_lm, data.frame(Year = c(2013, 2015, 2020, 2023)))
```



```{r}
# make a linear model with time to predict country temperatures
# first check correlation
temps_usa_y <- temps_by_country_y |> filter(ISOName == "USA")
cor(temps_usa_y$Year, temps_usa_y$YearAvgF)
# nice correlation, let's look for outliers

ggplot(temps_usa_y) + geom_boxplot(aes(y = YearAvgF)) + labs(y = "Yearly Average Temperature (F)", title = "Yearly Average Temperature in the US")
# not too bad on outliers, let's keep going

# look for normality
ggplot(temps_usa_y) + geom_histogram(aes(x = YearAvgF)) + labs(x = "Yearly Average Temperature (F)", y = "Count", title = "Yearly Average Temperature Distribution in the US")
# pretty normal, let's make the linear model now
```

```{r}
usa_lm <- lm(YearAvgF ~ Year, data = temps_usa_y)
summary(usa_lm)
```

```{r}
eq_usa = paste0("US Yearly Averages vs. Temperature Model: y = ", round(usa_lm$coefficients[2],2),
            "*x + ", round(usa_lm$coefficients[1],2))

temps_usa_y |>
  ggplot(aes(x = Year, y = YearAvgF)) + geom_point(size = 2) +
  xlab("Year") +
  ylab("US Yearly Averages") +
  # add an arbitrary line with slope and intercept
  geom_abline(intercept = usa_lm$coefficients[1], # extract first coefficient from model
              slope = usa_lm$coefficients[2], # ditto
              linetype = "dashed",
              color = "blue",
              linewidth = 1.5) +
  ggtitle(eq_usa) +
  theme(plot.title = element_text(color="blue", size=14, face="bold"))
```



```{r}
# make a linear model with latitude and time predicting city's temperatures
# first check correlation

temps_city_lat <- temps_by_city_y |> select(Year, LatNumeric, YearAvgF) |> ungroup()
cor(temps_city_lat$Year, temps_city_lat$YearAvgF)
cor(temps_city_lat$LatNumeric, temps_city_lat$YearAvgF)
# nice correlation, let's look for outliers

ggplot(temps_city_lat) + geom_boxplot(aes(y = LatNumeric)) + labs(y = "Yearly Average (F)", title = "Yearly Average Temperatures by City")
# not too bad on outliers, let's keep going

# look for normality
ggplot(temps_city_lat) + geom_histogram(aes(x = YearAvgF)) + labs(x = "Yearly Average (F)", y = "Count", title = "Yearly Average Temperature Distribution")

ggplot(temps_city_lat) + geom_point(aes(x = LatNumeric, y = YearAvgF)) + labs(x = "Latitude Distance from Equator", y = "Yearly Average Temperature by City", title = "Latitude vs. Temperature")
# pretty normal, let's make the linear model now
```


```{r}
lat_lm <- lm(YearAvgF ~ LatNumeric, data = temps_city_lat)
summary(lat_lm)
```


```{r}
eq_lat = paste0("Yearly Average Temperature by City (F) vs. Latitude Model: y = ", round(lat_lm$coefficients[2],2),
            "*x + ", round(lat_lm$coefficients[1],2))

temps_city_lat |>
  ggplot(aes(x = LatNumeric, y = YearAvgF)) + geom_point(size = 2) +
  xlab("Latitude Distance from Equator") +
  ylab("Yearly Average Temperature by City (F)") +
  # add an arbitrary line with slope and intercept
  geom_abline(intercept = lat_lm$coefficients[1], # extract first coefficient from model
              slope = lat_lm$coefficients[2], # ditto
              linetype = "dashed",
              color = "blue",
              linewidth = 1.5) +
  ggtitle(eq_lat) +
  theme(plot.title = element_text(color="blue", size=14, face="bold"))
```






